\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-02-23 paper/si.tex
milestone in writing

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the update
phase an individual adopts the strategy of another individual based on their
``updating payoffs''. In Section~\ref{section:updating_payoffs} we
present/describe the conventional approach for calculating updating payoffs, as
well as, our newly introduced approach.

@@

In the \textbf{mutation phase} one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\).
In the \textbf{game phase} individuals are randomly matched with other
individuals in the population, and they engage in a repeated game where each
subsequent turn occurs with a fixed probability \(\delta\). At each turn the
individuals decide on an action based on their strategies.
In repeated games there are infinite many strategies, however, it is commonly
assumed that individuals can only choose strategies from a restricted set. One
such set is that of reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently.

In the \textbf{update stage} where two individuals are randomly selected. From
the two individuals, one  serves as the `learner' and the other as the `role
model'. The learner adopts the role model's strategy with a probability \(\rho\)
given by,

@@

\begin{equation}\label{eq:appendix_fixation_probability}
    \varphi = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}},
\end{equation}

@@

\begin{algorithm}[!htbp]
  \SetAlgoLined $N \leftarrow$ population size\; $k \leftarrow 1$\; resident
   $\leftarrow$ starting resident\; \While{ $t <$ maximum number of steps}{mutant
   $\leftarrow$ random: $\{\emptyset \}\ \rightarrow R^n$\; fixation probability
   $\leftarrow \varphi $\; \If{$\varphi >$ random: $i \rightarrow [0,
   1]$}{resident $\leftarrow$ mutant;}} \caption{Evolutionary
   process}\label{algorithm:pairwise_comparison}
\end{algorithm}

Most of the results we present in this work consider the case of low mutation,
however, we have also verified that the main result holds in the case of
high mutation rates. %ToDo Add reference to section with mu != 0.

@@

The updating payoffs depend on the interactions of the individuals at the game
phase. In this work we assume that in the game phase individuals are matched in
pairs and that they participate in a repeated 2 persons donation game. In the
donation game there are two actions: cooperation (\(C\)) and defection (\(D\)).
By cooperating a player provides a benefit \(b\) to the other player at their
cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn are,

@@

In the following subsections we present several approaches for calculating
the updating payoffs. Initially, we discuss the conventional approach of the
expected payoffs and afterwards we present our newly introduced approach.

\subsection{Updating Payoffs based on the expected payoffs}

The expected payoffs are the conventional payoffs used in the updating stage.
They are defined as the mean payoff of an individual in a well-mixed population
that engages in an infinitely repeated games with all other population members. In
an infinitely repeated game the payoff of a reactive strategy can explicitly be
calculated using a markovian approach~\cite{}. Namely, assume two reactive
strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$, their play can
be defined as a Markov process with the transition matrix \(M\),

@@

Note that \(y_1 \text{ and } y_2\) are omitted since they have no effect in the
long run of the game. The stationary vector \(\mathbf{v}\),
which is the solution to \(\mathbf{v} M = \mathbf{v}\), combined with the payoff vector
\(\mathbf{u}\), yields the game stage outcome for each strategy,

@@

@@

\subsubsection*{Simulation Results based on the expected payoffs}

We can simulate the evolutionary process described by
Algorithm~\ref{algorithm:pairwise_comparison} when individuals use reactive
strategies and update their strategies based on their expected payoffs. We
performed two independent runs of the process where we differed the benefit of
cooperation \(b\). The results are shown in
Figure~\ref{fig:expected_payoffs_results}.

It is observed that a higher value of benefit results in a more cooperative
population. For a low benefit the resident population cooperates on average 52\%
of the time. For a a high benefit the average cooperation increases to 98\%. For
both a low and a high benefit the resident population consists either of
defectors or conditional cooperators. A conditional cooperator, otherwise
known as a generous tit for tat (GTFT) strategy, always cooperates if the
co-player cooperated ($p\approx1$) and cooperates with a probability \(q\) if
the co-player defected. The conditional cooperator strategy adopted by the
population differs between the two simulations. A higher value of benefit
results to a higher tolerance to defection.

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=.70\textwidth]{static/expected_payoffs_donation_game.pdf}
    \caption{\textbf{Evolutionary dynamics under expected payoffs with
    low (left) and high (right) benefit.} We run the simulations for \(T =
    10^7\) time steps. For each time step, we have recorded the current resident
    population \((y, p, q)\). Since in the case of the expected payoffs the
    individuals interact for an infinite number of times \(\delta \rightarrow
    1\), we do not report the players' initial cooperation probability \(y\).
    The graphs show how often the resident population chooses each combination
    \((p, q)\) of conditional cooperation probabilities in the subsequent
    rounds. In both cases players update based on their expected payoffs (A) In
    the case where \(b=3\), the resident population either consists of defectors
    (with \(p \approx q \approx 0\)) or of conditional cooperators for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{3}=0.7\). (B) In the case where
    \(b=10\), the resident population typically applies a strategy for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{10}=0.9\). Parameters: \(N =100, c=1,
    \beta=1\).
    }\label{fig:expected_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the expected payoffs}

We can explicitly calculate the value of \(q\) which evolves. Namely, we will
calculate how easily a single ALLD mutant can invade into a resident population
with strategy GTFT. In that case, GTFT\(= (1, 1, q)\), ALLD\(= (0, 0, 0)\), and
\(k = 1\).

When two GTFT players interact in an infinitely repeated game, their respective
probabilities for each of the four outcomes in the last round simplify to,

@@

The aim of this work is to explore the effect that the updating payoffs have on
the cooperation rate of the evolved population. To this end we introduce
a new method calculating the updating payoffs. Compared to the literature
where a infinite number of rounds and a well mixed population are assumed,
we constrain the number of turns and number of turns.

Namely, in the expected payoffs the payoff of a pair depends on the average
payoff they received over an infinite number of turns. In our approach, the
payoff of a pair depends on the average payoffs they received in the last \(m\)
turns. Furthermore, in expected payoffs it is assumed that a player interacts
with every member of the population, whereas in our approach with consider one
\(n\) interactions of the player with other members. Here we present results for
four cases; (\(m=1\) and \(n=1\)), (\(m=2\) and \(n=1\)), (\(m=1\) and \(n=2\)),
and (\(m=2\) and \(n=2\)).

\subsubsection{Updating Payoffs based on the last round payoff of one interaction (\(m=1\) and \(n=1\))}\label{section:m_one_n_one}

@@

At each step of the evolutionary process we choose a role model and a learner to
update the population. We consider the case where both the role model and the
learner estimate their fitness after interacting with a single member of the
population. There are five possible pairings at each step. They interact with
other with a probability \(\frac{1}{N - 1}\), and thus they do not interact with
other with a probability \(1 - \frac{1}{N - 1}\). In the latter case, each of
them can interact with either a mutant or a resident. Both of them interact with
a mutant with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact
with a resident with a probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last
two possible pairings are that either of them interacts with a resident whilst
the other interacts with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.

Given the possible pairings and Proposition~\ref{proposition:last_round}, we
define the probability that the respective last round payoffs of two players
\(s_1, s_2\) are given by $u_1$ and $u_2$ as,

@@

In this expression, $\frac{(N\!-\!k)}{N}$ is the probability that the randomly
chosen learner is a resident, and $\frac{k}{N}$ is the probability that the role
model is a mutant. The sum corresponds to the total probability that the learner
adopts the role model's strategy over all possible payoffs $u_R$ and $u_M$ that
the two player may have received in their respective last rounds. We use
$x(u_R,u_M)$ to denote the probability that the randomly chosen resident
obtained a payoff of $u_R$ in the last round of his respective game, and that
the mutant obtained a payoff of $u_M$.

\subsubsection*{Simulation Results based on the last round payoff of one interaction}

We simulate the evolutionary process given that individuals now use the the last
round payoff of one interaction to update their strategies. The results are
shown in Figure~\ref{fig:one_interaction_last_round_payoffs_results}. Similarly
to the results of the expected payoffs, a higher benefit results in a more
cooperative population. However we note that the cooperation rate of the evolved
populations are lower compared to the case of expected payoffs, and that the
cooperation rate increases less between the two simulations. The evolving both
consist of defectors and conditional cooperators. The evolved \(q\)s are smaller
which also results to a less cooperative population. In between the two
simulations \(q \approx \frac{1}{2}\). In the invasion analysis we will
demonstrate that in the case of the the last round payoff of one interaction, a
conditional player needs to have \(q \leq \frac{1}{2}\) to avoid being invaded
by a defector. In both simulations \(q \leq \frac{1}{2}\), the higher benefit
pushes the evolved population closer to the bounty.

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/one_interaction_last_round_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under one interaction last round
  payoffs with low (left) and high (right) benefit.} The cooperation rates
  in both cases are less compared to the expected payoffs. Moreover, increase is
  less when the benefit is increased to ten. In the cases of low and
  high benefit case the resident population either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1\).
  }\label{fig:one_interaction_last_round_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the last round payoff of one interaction}

We once again calculate how easily a single ALLD mutant can invade into a
resident population with strategy GTFT. When two GTFT players interact in the
game, their respective probabilities for each of the four outcomes in the last
round simplify to,

@@

Here, in the updating phase we choose one role model and one learner, however
now, we consider the case where both the role model and the learner estimate
their updating payoff after interacting with two members of the population.
There are stages in the pairing process and there are twenty four possible
pairings.

In the first stage the role model and the learner are matched together with a
probability $\frac{1}{N-1}$ or not with a probability $(1 - \frac{1}{N-1})$. If
they were matched then in the second stage of pairings there are only four
possible outcomes; both of them interact with a mutant with a probability
$\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two possible pairings
are that either of them interacts with a resident whilst the other interacts
with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.

In the later case, where they were not matched in the first stage, there are
four possible outcomes; both of them interact with a mutant and both interact
with a resident, either of them interacts with a resident whilst the other
interacts with a mutant. For each of the above pairings of the first stage there
are five possible; they interact with each other, both of them interact with a
mutant and both interact with a resident, either of them interacts with a
resident whilst the other interacts with a mutant.

The new possible pairings change how we define the probability that the
respective last round payoffs of two players \(s_1, s_2\) are given by \(u_1\)
and \(u_2\). The probability denoted as \(\bar{x}(u_1,u_2)\) is given by,

\begin{equation}
  \resizebox{.99\textwidth}{!}
  {%
$
  \setlength{\arraycolsep}{1pt}
  \begin{array}{llrl}
  x(u_1,u_2)	& = \displaystyle \frac{1}{N\!-\!1}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} \cdot A + \left(1\!-\!\frac{1}{N\!-\!1}\right) \left[ \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_2) _{u_2}(s_2,s_2)  \frac{\left(k - 2\right) \left(k - 1\right)}{(N\!-\!2)(N\!-\!3)}
  \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [B_1 + B_2 + B_3 + B_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_2) _{u_2}(s_2,s_1)  \frac{\left(k - 1\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [C_1 + C_2 + C_3 + C_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_1) _{u_2}(s_2,s_2)  \frac{\left(k - 1\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [D_1 + D_2 + D_3 + D_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_1) _{u_2}(s_2,s_1)  \frac{\left(N - k - 2\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [E_1 + E_2 + E_3 + E_4]\right) \right]
  \end{array} $}
\end{equation}

@@

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the first stage of pairing,
followed by them being paired with another member of the population on the
second stage. The second terms corresponds to the case that the learner and the
role model interact with a mutant with a probability ($\frac{\left(k - 2\right)
\left(k - 1\right)}{(N\!-\!2)(N\!-\!3)}$). In the seconds stage, they can either
interact with each other $\frac{1}{N - 2}$ or not $(1 - \frac{1}{N - 2})$. If
they do not interact with each other, then each of the following can happen:
both of them interact with a mutant with a probability
$\frac{(k-3)(k-2)}{(N-4)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-3)(N-4)}$. The last two possible pairings
are that either of them interacts with a resident whilst the other interacts
with a mutant, and this happens with a probability
$\frac{(N-k-2)(k-1)}{(N-4)(N-3)}$. The rest of the cases follow the same
pattern.

@@

\begin{align}
\lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} \bar{x}(u_{R},u_{M})\cdot \rho(u_{R},u_{M}), \\
\lambda^-_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} \bar{x}(u_{R},u_{M})\cdot \rho(u_{M},u_{R}).
\end{align}

\subsubsection*{Simulation Results based on the last round payoff of two interactions}

We simulate the evolutionary process when individuals update their strategies
based on the last round payoffs from two interactions with other members of the
population, Figure~\ref{fig:one_interaction_last_round_payoffs_results}. For
both a low and a high benefit value the average cooperation rate has increased
compared to the the last round against a single other member of the population.
In the case of low benefit the evolved population's cooperation rate is almost
the same as in the case of expected payoffs. However, the cooperation rate does
not increase as much as in for the expected payoffs for \(b=10\).

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/two_interactions_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under two interactions last round
  payoffs with low (left) and high (right) benefit.} In the cases of low and
  high benefit case the resident population either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{\sqrt{2}}{2}\). Parameters: \(N =100, c=1, \beta=1\).
  }\label{fig:two_interactions_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the last round payoff of two interactions}

In a similar fashion we can calculate the threshold of \(q\) for which a GTFT
player can not be invaded by an ALLD mutant. The ratio of transition probabilities
is given by,

@@

In the limit of strong selection \(\beta \rightarrow \infty\) and large
populations \(N \rightarrow \infty \), we obtain

>>>>

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the update
phase an individual adopts the strategy of another individual based on their
``updating payoffs''. In Section~\ref{section:updating_payoffs} we
describe the conventional approach for calculating updating payoffs and we
present our new approaches.

@@

In the \textbf{mutation phase} one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\).
In the \textbf{game phase} individuals are randomly matched with other
individuals in the population and they engage in a repeated game where each
subsequent turn occurs with a fixed probability \(\delta\). At each turn the
individuals decide on an action based on their strategies.
In repeated games there are infinitely many strategies, however, it is commonly
assumed that individuals can only choose strategies from a restricted set. One
such set is that of reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently.

In the \textbf{update stage} two individuals are randomly selected. From
the two individuals, one  serves as the `learner' and the other as the `role
model'. The learner adopts the role model's strategy with a probability \(\rho\)
given by,

@@

\begin{equation}\label{eq:appendix_fixation_probability}
    \varphi_{M} = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}},
\end{equation}

@@

\begin{algorithm}[!htbp]
  \SetAlgoLined
  $N \leftarrow$ population size\;
  $k \leftarrow 1$\; resident
   $\leftarrow$ starting resident\;
   \While{ $t <$ maximum number of steps}
   {mutant $\leftarrow$ random strategy\;
   fixation probability $\leftarrow \varphi_M $\;
   \If{$\varphi_M >$ random: $i \rightarrow [0,1]$}
   {resident $\leftarrow$ mutant;}}
   \caption{Evolutionary process}\label{algorithm:pairwise_comparison}
\end{algorithm}

Most of the results we present in this work consider the case of low mutation,
however, we have also verified that the main result holds in the case of
high mutation rates (Section~\ref{section:mutation}).

@@

The updating payoffs depend on the interactions of the individuals at the game
phase. In this work we assume that in the game phase individuals are matched in
pairs and that they participate in a repeated 2 person donation game. In the
donation game there are two actions: cooperation (\(C\)) and defection (\(D\)).
By cooperating a player provides a benefit \(b\) to the other player at their
cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn are,

@@

In the following subsections we present several approaches for calculating
the updating payoffs. Initially, we discuss the conventional approach of the
expected payoffs and afterwards we present our new approach.

\subsection{Updating Payoffs based on the expected payoffs}\label{section:perfect_memory}

The expected payoffs are the conventional payoffs used in the updating stage. We
refer to this approach as the perfect memory approach. They are defined as the
mean payoff of an individual in a well-mixed population that engages in an
infinitely repeated games with all other population members. In an infinitely
repeated game the payoff of a reactive strategy can explicitly be calculated
using a markovian approach~\cite{nowak:APC:1989}. Namely, assume two reactive
strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$, their play can
be defined as a Markov process with the transition matrix \(M\),

@@

The stationary vector \(\mathbf{v}(s_1,s_2)\),
which is the solution to \(\mathbf{v}(s_1,s_2) M = \mathbf{v}(s_1,s_2)\), combined with the payoff vector
\(\mathbf{u}\), yields the game stage outcome for each strategy,

@@

\(\frac{N\!-\!k\!-\!1}{N-1}\) is the probability that a resident meets another
resident and \(\frac{k}{N-1}\) is the probability that a resident meets a
mutant. Likewise, \(\frac{N-k}{N-1}\) is the probability that a mutant interacts
with a resident and \(\frac{k-1}{N-1}\) the probability that the mutants
interacts with a mutant.

@@

\subsubsection*{Simulation Results based on Expected Payoffs}

We simulate the evolutionary process, as described in
Algorithm~\ref{algorithm:pairwise_comparison}, when individuals use reactive
strategies and update their strategies based on their expected payoffs. We
performed two independent runs of the process where we differed the benefit of
cooperation \(b\). The results are shown in
Figure~\ref{fig:expected_payoffs_results}. We observed that a higher value of
benefit results in a more cooperative population. For a low benefit the resident
population cooperates on average 52\% of the time. For a high benefit the
average cooperation increases to 98\%. For a low benefit the resident population
consists either of defectors or conditional cooperators, and for a high value of
benefit the population predominantly consists of conditional cooperators.A
conditional cooperator, otherwise known as a generous tit for tat (GTFT)
player, always cooperates if the co-player cooperated ($p\approx1$) and
cooperates with a probability \(q\) if the co-player defected. The conditional
cooperator strategy adopted by the population differs between the two
simulations, namely, a higher value of benefit results in higher values
of \(q\).

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=.70\textwidth]{static/expected_payoffs_donation_game.pdf}
    \caption{\textbf{Evolutionary dynamics under expected payoffs with
    low (left) and high (right) benefit.} We run the simulations for \(T =
    10^7\) time steps. For each time step we recorded the current resident
    population \((y, p, q)\). Since \(\delta \rightarrow
    1\) we do not report the players' initial cooperation probability \(y\).
    The graphs show how often the resident population chooses each combination
    \((p, q)\) of conditional cooperation probabilities in the subsequent
    rounds. In both cases players update based on their expected payoffs.
    ({\bf Low Benefit}) In
    the case where \(b=3\), the resident population either consists of defectors
    (with \(p \approx q \approx 0\)) or of conditional cooperators for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{3}=0.7\). ({\bf High Benefit}) In the case where
    \(b=10\), the resident population typically applies a strategy for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{10}=0.9\). Parameters: \(N =100, c=1,
    \beta=1\).
    }\label{fig:expected_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on Expected Payoffs}

We can explicitly calculate the value of \(q\) which evolves. Namely, we will
calculate how easily a single defecting mutant can invade into a resident
population of GTFT players. Let GTFT\(= (1, 1, q)\), ALLD\(= (0, 0, 0)\), and
\(k = 1\). When two GTFT players interact in an infinitely repeated game, their
respective probabilities for each of the four outcomes in the last round
simplify to,

@@

The aim of this work is to explore the effect of the updating payoffs on the
evolution of cooperation. To this end we introduce a new method for calculating
the updating payoffs and we compare the results of the new method to those of
the expected payoffs. In our approach we do not assume that an individual has a
perfect memory. Instead an individual remembers only a limited number of
interactions and limited outcomes from each interaction. In the expected payoffs
case the payoff of a pair depends on the average payoff they received over an
infinite number of turns. In our approach, the payoff of a pair depends on the
average payoffs they received in the last \(m\) turns. Furthermore, in expected
payoffs it is assumed that a player interacts with every member of the
population, whereas in our approach a player has \(n\) interactions.  Here we
present results for four cases; (\(m=1\) and \(n=1\)), (\(m=2\) and \(n=1\)),
(\(m=1\) and \(n=2\)), and (\(m=2\) and \(n=2\)).

\subsubsection{Updating Payoffs based on the Last Round Payoff of One Interaction (\(m=1\) and \(n=1\))}\label{section:m_one_n_one}

@@

At each step of the evolutionary process we choose a role model and a learner to
update the population. In this case both the role model and the learner estimate
their fitness after interacting with a single member of the population, and so
there are five possible pairings at each step. They interact with each other with a
probability \(\frac{1}{N - 1}\), and they do not interact with other with a
probability \(1 - \frac{1}{N - 1}\). In the latter case, each of them can
interact with either a mutant or a resident. Both of them interact with a mutant
with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact with a
resident with a probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two
possible pairings are that either of them interacts with a resident whilst the
other interacts with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$. Given the possible pairings and
Proposition~\ref{proposition:last_round}, we define the probability that the
respective last round payoffs of two players \(s_1, s_2\) are given by $u_1$ and
$u_2$ as,

@@

In this expression, $\frac{(N\!-\!k)}{N}$ is the probability that the randomly
chosen learner is a resident, and $\frac{k}{N}$ is the probability that the role
model is a mutant. The sum corresponds to the total probability that the learner
adopts the role model's strategy over all possible payoffs $u_R$ and $u_M$ that
the two players may have received in their respective last rounds. We use
$x(u_R,u_M)$ to denote the probability that the randomly chosen resident
obtained a payoff of $u_R$ in the last round of his respective game, and that
the mutant obtained a payoff of $u_M$.

\subsubsection*{Simulation Results based on the Last Round Payoff of One Interaction}

We simulate the evolutionary process given that individuals now use the last
round payoff of one interaction to update their strategies. The results are
shown in Figure~\ref{fig:one_interaction_last_round_payoffs_results}. Similarly
to the results of the expected payoffs, a higher benefit results in a more
cooperative population. However, we note that the cooperation rate of the
evolved populations is lower compared to the case of expected payoffs, and that
the increase of the cooperation rate between the two simulations is smaller. The
residents population in both cases consist of defectors and conditional
cooperators with \(p \approx 1\) and  \(q \approx \frac{1}{2}\). Thus, the
evolved \(q\)s are smaller which is the reason why the population is less
cooperative. In the invasion analysis we demonstrate that in the case of the
last round payoff of one interaction, a conditional player needs to have \(q
\leq \frac{1}{2}\) to avoid being invaded by a defector regardless of the
benefit of cooperation.

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/one_interaction_last_round_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under one interaction last round
  payoffs with low (left) and high (right) benefit.} The cooperation rate
  regardless of the benefit of cooperation is less compared to that of the
  expected payoffs. The increase of cooperation between the low benefit case and
  the high benefit case is also smaller. In both cases the resident population
  either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1, T=10^{7}\).
  }\label{fig:one_interaction_last_round_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the Last Round Payoff of One Interaction}

We once again calculate how easily a single ALLD mutant can invade into a
resident population of GTFT player. When two GTFT players interact in the game,
their respective probabilities for each of the four outcomes in the last round
simplify to,

@@

Here we consider the case where both the role model and the learner estimate
their updating payoff after interacting with two members of the population.
There are two stages of matching and there are twenty four possible pairs. In
the first stage the role model and the learner are matched together with a
probability $\frac{1}{N-1}$ or not with a probability $(1 - \frac{1}{N-1})$. If
they were matched together then in the second stage there are only four possible
outcomes; both of them interact with a mutant with a probability
$\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two possible pairs are
that either of them interacts with a resident whilst the other interacts with a
mutant, and this happens with a probability $\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.
In the later case, where they were not matched in the first stage, there are
four possible outcomes; both of them interact with a mutant and both interact
with a resident, either of them interacts with a resident whilst the other
interacts with a mutant. For each of the above pairs of the first stage there
are five possible pairs in the second stage; they interact with each other, both
of them interact with a mutant and both interact with a resident, either of them
interacts with a resident whilst the other interacts with a mutant.

The new possible pairs change how we define the probability that the
respective last round payoffs of two players \(s_1, s_2\) are given by \(u_1\)
and \(u_2\). The new probability denoted as \(\tilde{x}(u_1,u_2)\) is given by,

\begin{equation}
  \resizebox{.99\textwidth}{!}
  {%
$
  \setlength{\arraycolsep}{1pt}
  \begin{array}{llrl}
  \tilde{x}(u_1,u_2)	& = \displaystyle \frac{1}{N\!-\!1}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} \cdot A + \left(1\!-\!\frac{1}{N\!-\!1}\right) \left[ \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_2) _{u_2}(s_2,s_2)  \frac{\left(k - 2\right) \left(k - 1\right)}{(N\!-\!2)(N\!-\!3)}
  \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [B_1 + B_2 + B_3 + B_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_2) _{u_2}(s_2,s_1)  \frac{\left(k - 1\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [C_1 + C_2 + C_3 + C_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_1) _{u_2}(s_2,s_2)  \frac{\left(k - 1\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [D_1 + D_2 + D_3 + D_4]\right) + \right. \\[0.5cm] 
  & \left. v_{u_1}(s_1,s_1) _{u_2}(s_2,s_1)  \frac{\left(N - k - 2\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)} \left(\frac{1}{N\!-\!2}\cdot v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F} + (1 - \frac{1}{N - 2}) [E_1 + E_2 + E_3 + E_4]\right) \right]
  \end{array} $}
\end{equation}

@@

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the first stage of pairing,
followed by them being paired with another member of the population on the
second stage. The second terms corresponds to the case that the learner and the
role model interact with a mutant with a probability ($\frac{\left(k - 2\right)
\left(k - 1\right)}{(N\!-\!2)(N\!-\!3)}$). In the seconds stage, they can either
interact with each other $\frac{1}{N - 2}$ or not $(1 - \frac{1}{N - 2})$. If
they do not interact with each other, then each of the following can happen:
both of them interact with a mutant with a probability
$\frac{(k-3)(k-2)}{(N-4)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-3)(N-4)}$. The last two possible pairings
are that either of them interacts with a resident whilst the other interacts
with a mutant, and this happens with a probability
$\frac{(N-k-2)(k-1)}{(N-4)(N-3)}$, and so on.

@@

\begin{align}\label{eq:ratio_limited}
\lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} \tilde{x}(u_{R},u_{M})\cdot \rho(u_{R},u_{M}), \\
\lambda^-_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} \tilde{x}(u_{R},u_{M})\cdot \rho(u_{M},u_{R}).
\end{align}

\subsubsection*{Simulation Results based on the Last Round Payoff of Two Interactions}

We simulate the evolutionary process when individuals update their strategies
based on the last round payoffs from two interactions with members of the
population, Figure~\ref{fig:one_interaction_last_round_payoffs_results}. For
both a low and a high benefit value the average cooperation rate has increased
compared to the last round against a single other member of the population. In
the case of low benefit the evolved population's cooperation rate is almost the
same as in the case of expected payoffs. However, the cooperation rate does not
increase as much as the expected payoffs for \(b=10\).

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/two_interactions_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under two interactions last round
  payoffs with low (left) and high (right) benefit.}
  The cooperation rates for both the simulations have increased and are similar
  to the case of expected payoffs. The increase of the cooperation rate between
  the two cases however is smaller. In both cases the resident population
  either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1, T=10^{7}\).
  }\label{fig:two_interactions_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the Last Round Payoff of Two Interactions}

In a similar fashion we can calculate the condition for which a population of
GTFT players can not be invaded by an ALLD mutant. The ratio of transition
probabilities is given by,

@@

In the limit of strong selection \(\beta \rightarrow \infty\) and large
populations \(N \rightarrow \infty \) we obtain the following cases,
